---
title: "Reproducible Research- Project 1"
author: "Carlos Martinez"
date: "26/8/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

## 1. Preparation, loading and preprocessing the data
### a).  Load the data 
```{r echo=TRUE}
data<-read.csv("activity.csv")
head(data)
# Note: there are NA values in the steps variable
```
### b).  Process/transform the data (if necessary) into a format suitable for your analysis
```{r, echo=TRUE}
dev_null <- Sys.setlocale("LC_TIME", "english")
initdata <- read.csv("activity.csv", colClasses=c("numeric", "Date", "numeric"))
```


```{r, echo=TRUE}
data <- data[!(is.na(data$steps)), ]
```

```{r, echo=TRUE}
# We will use the aggregate () function.
totalstep <- aggregate(steps ~ date, data, sum)
head(totalstep)
# Note: there is not NA values in the steps variable
```
### 1.1 Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
totalstep <- with(data, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
head(totalstep)
names(totalstep) <- c("date", "steps")
color <- colorRampPalette(c("skyblue", "darkblue", "skyblue"))
hist(totalstep$steps, xlab=" Steps per day", ylab=" Number of Days ", main=" Histogram of the total number of steps taken each day ", col = color(32), ylim = c(0,20), breaks = seq(0,25000, by=2500))
```

### 1.2. Calculate and report the mean and median total number of steps taken per day 
```{r, echo=TRUE}
library(dplyr)
statistics <- summarise(totalstep, mean=mean(totalstep $steps),
                               median=median(totalstep $steps))
statistics
# We report the mean and median: Mean is 10766.19 and Median value 10765
```


### 2.1.  Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=TRUE}
library(ggplot2)
prom<- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=prom, aes(x=interval, y=steps)) +
    geom_path(colour = " darkblue", lwd=0.5)+
    ggtitle("Time Series: Averaged across all days / 5-minute interval")+
    xlab("5-minute interval") +
    ylab("Average number of steps taken")
```

### 2.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}
prom[which.max(prom[,2]),1]
#Maximum number of steps is the interval 835
```


```{r, echo=TRUE}
prom[which.max(prom[,2]),1]
#Maximum number of steps is the interval 835
```

### 3.1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo=TRUE}
datos<-read.csv("activity.csv")
colSums(is.na(datos)) 
# 2304 NA values contained in the steps variable, The VIM package helps us to see the proportion of NAs in the steps variable.
```

```{r, echo=TRUE}
library(VIM)
library(mice)
aggr(datos,numbers=T,sortVar=T)
# 13.11% of data in the steps variable are NAs
```


### 3.2 Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r, echo=TRUE}
# Imputation with the mean (37.3826) across all days, we use the package MICE
datos<-data.frame(datos)
columns <- c("steps","date","interval")
imputed_data <- mice(datos[,names(datos) %in% columns],m = 1, maxit = 1, method = "mean",seed = 2018,print=F)
```

```{r, echo=TRUE}
# NOTE: we can also impute the data with the mean with the following function:
meanStepsInterval <- aggregate(steps ~ interval, data, mean)
head(meanStepsInterval)
imputedData <- initdata
for(x in 1:17568) {
    if(is.na(imputedData[x, 1])==TRUE) {
        imputedData[x, 1] <- meanStepsInterval[meanStepsInterval$interval %in% imputedData[x, 3], 2]
    }
}
head(imputedData)
```


### 3.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE}
complete.data <- mice::complete(imputed_data)
complete<-aggregate(steps~date,complete.data,sum)
```
### 3.4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median

#### 3.4.1. Calculate the mean and median total number of steps per day after imputation
```{r, echo=TRUE}
statistics1 <- summarise(complete, mean=mean(complete $steps),
                               median=median(complete$steps))
statistics1
```
#### 3.4.2. Do these values differ from the estimates from the first part of the assignment? 

##### Answer: the means are equal the medians differ by 1.19

#### 3.4.3. What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r, echo=TRUE}
library(RColorBrewer)
darkcols <- brewer.pal(8, "Dark2")
par(mfrow = c(1, 2))
hist(totalstep $steps, breaks=20, , xlab=" Steps per day", ylab=" Number of Days ",
          col = darkcols, ylim=c(0, 20), main=NULL)
hist(complete$steps, breaks=20, , xlab=" Steps per day", ylab=" Number of Days ",
               col = darkcols, ylim=c(0, 20), main=NULL)
mtext("Histograms of the total number of steps taken each day Without & With imputed values", adj=1)
```


##### Answer:  Regarding the mean and median there is no considerable impact, but we can observe in the histograms that the frequency increases in the imputed data.


### 4.1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
"English_United States.1252"
daysData <- imputedData
daysData$days <- weekdays(daysData$date)
daysData$weekday <- as.character(rep(0, times=17568))
for(x in 1:17568) {
    if(daysData[x, 4] %in% c("Saturday", "Sunday")) {
        daysData[x, 5] <- "weekend"
    } else {
        daysData[x, 5] <- "weekday"
    }
}
daysData$weekday <- factor(daysData$weekday)
weekdayData <- daysData[daysData$weekday=="weekday", ]
weekendData <- daysData[daysData$weekday=="weekend", ]
weekdayMean <- aggregate(steps ~ interval, weekdayData, mean)
weekendMean <- aggregate(steps ~ interval, weekendData, mean)
```

### 4.2. Make a panel plot containing a time series plot

```{r, echo=TRUE}
par(mfrow=c(2, 1), mar=c(4, 4.1, 3, 2.1))
plot(weekdayMean$interval, weekdayMean$steps, type="l",
     main="Average Steps Taken per Interval for Weekdays",
     xlab="Intervals", ylab="Number of Steps",
     col="darkred", lwd=1.5, ylim=c(0, 230))
plot(weekendMean$interval, weekendMean$steps, type="l",
     main=" Average Steps Taken per Interval for Weekends",
     xlab="Intervals", ylab="Number of Steps",
     col="blue", lwd=1.5, ylim=c(0, 230))
```



